Equivalence Hypergraphs:\\ Semantics of E-Graphs and their Generalization to Monoidal Theories

### 2023-10-18 sections/introduction.tex
update intro

Note how the natural level of generality for our string diagrams is not Cartesian, but monoidal: the generators of a Cartesian category are required in order to embed e-graphs, but are orthogonal to the addition of formal sums. Thus we generalize e-graphs from algebraic to monoidal theories.

@@

Quotienting e-hypergraphs by the induced equivalence gives rise to a sound and complete model of semilattice enriched categories.  That is, we exhibit a translation functor $\llbracket-\rrbracket$ from the free semilattice enriched SMC over a symmetric monoidal theory to our (quotiented) category of e-hypergraphs such that 
morphisms $f = g$ if and only if there exists a sequence of DPO-rewrites (each induced by a structural equality or the theory) between their translations: $\llbracket f \rrbracket \rightsquigarrow \llbracket_{DPO} g \rrbracket$.

@@

Having defined e-hypergraphs and shown them to correctly model rewriting string diagrams for SMCs, equipped with a notion of "equivalence class", we note that the naive implementation of DPO-rewriting is inefficient: to find a redex within an e-hypergraph involves finding a structurally equivalent e-hypergraph which contains the redex as a subgraph. In general, this involves unsharing the e-box structure before a redex becomes available. Thus, we define a notion of e-matching for e-hypergraphs: that is, to find redexes working modulo the e-structure. In particular, we wish to locate the smallest subgraph $G' \subset G$ "containing" the redex $L$. Given this, the analogue of e-rewriting $L \to R$ is simple to define, due to its non-destructive nature: we rewrite $G' \to G' + R$ in $G$.

[Example diagrams]

\subsection*{Future Applications}

Example diagrams for circuits, trace (loop unrolling, PEGs), LC, ZX here.

>>>>

Note how the natural level of generality for our string diagrams is not Cartesian, but monoidal: the generators of a Cartesian category are required in order to embed e-graphs, but are orthogonal to the addition of formal sums. Thus we generalize e-graphs from algebraic to monoidal theories, giving rise to a  host of new potential applications.

@@

Quotienting e-hypergraphs by the induced equivalence gives rise to a sound and complete model of semilattice enriched categories.  That is, we exhibit a translation functor $\llbracket-\rrbracket$ from the free semilattice enriched SMC over a symmetric monoidal theory to our (quotiented) category of e-hypergraphs such that 
morphisms $f = g$ if and only if there exists a sequence of DPO-rewrites (each induced by a structural equality or the theory) between their translations: $\llbracket f \rrbracket \rightsquigarrow_{DPO} \llbracket g \rrbracket$.

@@

Having defined e-hypergraphs and shown them to correctly model rewriting string diagrams for SMCs, equipped with a notion of "equivalence class", we note that the naive implementation of DPO-rewriting is inefficient: to find a redex within an e-hypergraph involves finding a structurally equivalent e-hypergraph which contains the redex as a subgraph. In general, this involves unsharing the e-box structure before a redex becomes available.  Thus, we define a notion of e-matching for e-hypergraphs: that is, to find redexes working modulo the e-structure. In particular, we wish to locate the smallest subgraph $G' \subset G$ "containing" (in an appropriate sense) the redex $L$. Given this, the analogue of e-rewriting $L \to R$ is simple to define, due to its non-destructive nature: we rewrite $G' \to G' + R$ in $G$.

\begin{figure}
\[
	\tikzfig{combinatorial_semantics/example-rewriting}
\]
\caption{Example application of e-rewriting for rewrite $(x*y)/z \to x*(y/z)$: find an appropriate minimal sub-e-hypergraph containing the redex $(x*y)/z$ and non-destructively add the reduct $x*(y/z)$.  REDO THIS}
\end{figure}

\subsection*{Future Work and Applications}

We have already alluded to the fact we deal with only \textit{acyclic} e-graphs in this paper.  Nevertheless,  within this setting we present a generalization from Cartesian to (symmetric) monoidal theories,  giving rise to a host of potential new domains of application.  Further,  the string diagrammatic and graph rewriting framework we set up is ameenable to extension, including lifting the acyclicity restriction.

Symmetric monoidal theories have received significant attention over recent years: the ZX-calculus can be used to characterize quantum processes, with applications to optimization of quantum circuits [cite]; similar theories also characterize signal flow diagrams [cite], digital circuits [cite], as well as having application to computational linguistics [cite].  In future work, we hope to apply our generalization of e-graphs to optimization problems in these domains (e.g. Figure \ref{})

The theories described above often have more structure than simply being symmetric monoidal: for example, they may contain comonoids, Frobenius algebras or traced structure (as in the case of e-graphs with cycles),  and this structure can be exploited to give more natural (in the sense of factoring out more equations) combinatorial representations.  In the above cases, such representations and their DPO-rewriting theory is well-studied [cite, cite, cite].  We consider that our development of hierarchical "e-box" structure within string diagrams and their associated (e-)hypergraphs ( categorically,  semilattice enrichment) is essentially orthogonal to these developments, and so ought to be easily combined to give notions of e-hypergraphs specifically suited to the optimization of these particular monoidal theories.

A particularly interesting extension that we intend to consider is the combination of our "e-structure" with (Cartesian) \textit{closed} categories, allowing the natural combination of e-graph optimization techniques with the $\lambda$-calculus.  The corresponding combinatorial representation would build on the recent work [reverse diff,FSCD].

Example diagrams for circuits (loop unrolling, PEGs)

\begin{figure}
\[
	\scalebox{0.75}{
	\tikzfig{combinatorial_semantics/example-e-graphs}
	}
\]
\caption{Example applications of e-hypergraphs.  Left, the ZX-calculus.  Right, the $\lambda$-calculus as represented by string diagrams for Cartesian closed categories.  }
\end{figure}
